---
title: "Test and Visualise Gene Enrichment with GSEA over MSigDB"
author: 'Stefano Mangiola '
date: '2021-07-12'
output:
  html_document:
    df_print: paged
categories: Case study
tags:
- tidytranscriptomics
- gene enrichment
- GSEA
- MSigDB
- differential expression
lastmod: '2021-07-12T12:29:10+10:00'
keywords: []
description: ''
comment: no
toc: no
autoCollapseToc: no
postMetaInFooter: no
hiddenFromHomePage: no
contentCopyright: no
reward: no
mathjax: no
mathjaxEnableSingleDollar: no
mathjaxEnableAutoNumber: no
hideHeaderAndFooter: no
flowchartDiagrams:
  enable: no
  options: ''
sequenceDiagrams:
  enable: no
  options: ''
slug: []
---

```{r message=FALSE, warning=FALSE}
# dataset
library(airway)

# tidyverse core packages
library(tibble)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(purrr)

# tidyverse-friendly packages
library(plotly)
library(ggrepel)
library(GGally)
library(tidybulk)
library(tidySummarizedExperiment) # we'll load this below to show what it can do
library(enrichplot)
library(patchwork)
```

tidySummarizedExperiment provides a bridge between Bioconductor [SummarizedExperiment](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html) [@morgan2020summarized] and the tidyverse [@wickham2019welcome]. It enables viewing the
Bioconductor *SummarizedExperiment* object as a tidyverse tibble, and provides SummarizedExperiment-compatible *dplyr*, *tidyr*, *ggplot* and *plotly* functions. This allows users to get the best of both Bioconductor and tidyverse worlds.


Here we will demonstrate performing a bulk RNA sequencing analysis using  *tidySummarizedExperiment* and *tidybulk*. We will use data from the *airway* package, which comes from the paper by [@himes2014rna]. It includes 8 samples from human airway smooth muscle cells, from 4 cell lines. For each cell line treated (with dexamethasone) and untreated (negative control) a sample has undergone RNA sequencing and gene counts have been generated.


```{r}
# load airway RNA sequencing data
data(airway)
# take a look
airway
```

## Full pipeline

In one modular step it is possible to go from raw counts to enriched pathways

```{r}
airway %>%
  
  # Annotate
  mutate(entrez = AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db,
                                        keys = feature,
                                        keytype = "ENSEMBL",
                                        column = "ENTREZID",
                                        multiVals = "first"
  )) %>% 

  # Filter
  keep_abundant(factor_of_interest = dex) %>%
  
  # Test differential gene transcript abundance
  test_differential_abundance(
    ~ dex + cell,
    method="edger_robust_likelihood_ratio", 
    test_above_log2_fold_change = 1,
  ) %>% 
  
  # Test gene enrichment
  filter(PValue   %>% is.na %>% `!`) %>%
  test_gene_rank(
    .entrez = entrez,
    .arrange_desc = logFC ,
    species="Homo sapiens",
    gene_sets = c("H", "C2", "C5")
  )
```

## Step-by-step

We'll set up the airway data for our RNA sequencing analysis. We'll create a column with shorter sample names and a column with entrez ID. We can get the entrez ID for these Ensembl gene ids using the Bioconductor annotation package for human, `org.Hs.eg.db`.


```{r}
# setup data workflow
counts <-
  airway %>%
  mutate(entrez = AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db,
                                        keys = feature,
                                        keytype = "ENSEMBL",
                                        column = "ENTREZID",
                                        multiVals = "first"
  )) 
# take a look
counts
```

We filter out lowly expressed genes using tidybulk `keep_abundant` or `identify_abundant`. These functions can use the *edgeR* `filterByExpr` function described in [@law2016rna] to automatically identify the genes with adequate abundance for differential expression testing.


```{r}
# Filtering counts
counts_abundant <- counts %>% 
      keep_abundant(factor_of_interest = dex) 

# take a look
counts_abundant
```

*tidybulk* integrates several popular methods for differential transcript abundance testing: the edgeR quasi-likelihood [@chen2016reads] (tidybulk default method), edgeR likelihood ratio [@mccarthy2012differential], limma-voom [@law2014voom] and DESeq2 [@love2014moderated]. A common question researchers have is which method to choose. With tidybulk, we can easily run multiple methods and compare.

We give `test_differential_abundance` our tidybulk counts object and a formula, specifying the column that contains our groups to be compared. If all our samples were from the same cell line, and there were no additional factors contributing variance, such as batch differences, we could use the formula `~ dex`. However, each treated and untreated sample is from a different cell line, so we add the cell line as an additional factor `~ dex + cell`.

```{r message=FALSE}
de_all <-
  counts_abundant %>%
  test_differential_abundance(
    ~ dex + cell,
    method="edger_robust_likelihood_ratio", 
    test_above_log2_fold_change = 1,
  ) 

```

Execute `GSEA` using `MSigDB`, `clusterProfiler` and `enrichplot` 

```{r}

de_all_gene_rank = 
  de_all %>%
  filter(PValue   %>% is.na %>% `!`) %>%
  test_gene_rank(
    .entrez = entrez,
    .arrange_desc = logFC ,
    species="Homo sapiens",
    gene_sets = c("H", "C2", "C5")
  )
```

Examine significantly enriched gene sets

```{r}
de_all_gene_rank %>%
  filter(gs_cat  == "C2" ) %>%
  dplyr::select(-fit) %>%
  unnest(test) %>% 
  filter(p.adjust < 0.05)

```

Visualise enrichment

```{r}
de_all_gene_rank_plots = 
  de_all_gene_rank  %>% 
    unnest(test) %>% 
  
  # Select top 10
    slice(1:10) %>% 
    mutate(plot = pmap(
       list(fit, ID, idx_for_plotting, p.adjust), 
       ~ enrichplot::gseaplot2(
           ..1, 
           geneSetID = ..3, 
           title = sprintf("%s \nadj pvalue %s", ..2, round(..4, 2)),
           base_size = 6, rel_heights = c(1.5, 0.5), subplots = c(1, 2)
       )  
  ))

# Visualise the first plot
de_all_gene_rank_plots %>% 
  pull(plot) %>% 
  wrap_plots()

```

List all citations used in this analysis
```{r}
  get_bibliography(de_all_gene_rank)
  
```

